\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2023-10-15 paper.tex
Minor tweaks and notes

In many settings, nodes are dated, i.e.\
each node $u\in N$ is associated with a time $\tau_u$.
However, the topological relationships
defined by the directed graph structure is sufficient for
many applications, and we will assume that node dates are not
known here.
% This should be obvious, but seems like it needs to be said
In practical settings, we will wish to associate additional
metadata such as sample identifiers or quality-control metrics
with nodes; it is therefore best to think of the
integers used here in the definition of a node as an \emph{identifier},
with which arbitrary additional information can be associated.

@@

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The pedigree relationship between the diploid individuals is
shown, along with their paths
of genetic inheritance and constituent genomes (labelled
with lowercase alphabetical letters
rather than integer identifiers to avoid confusion with genomic
intervals).
Thus, individual $D_1$ is composed
of genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents, $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the genomes contained within a parent. In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
genome \noderef{a} is the recombinant of
$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
Specifically, genome \noderef{a} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations, the corresponding gARG, and finally the corresponding
sequence of local trees along the genome.
The local trees represent the genealogy of genomes \noderef{a}--\noderef{d},
which changes between local trees due to recombination.
The local trees span the three genome regions delineated
by the two recombination breakpoints that gave rise to these genomes.
Following the above definitions, corresponding data encoding for
this gARG is shown in Tab.~\ref{tab-gARG-data} (Appendix~\ref{sec-gARG-data}).

These definitions follow naturally from the outcome of events
associated with the inheritance of DNA between generations.
As such,
there are some important consequences to these definitions.
Firstly, there is
no limitation on a mating system, ploidy, or age structure in a gARG.
By making the
basic unit an individual's genomes,
any pattern of inheritance among mixed ploidy (e.g.\ haplodiploid
species) can be accommodated.
Secondly, because we record the intervals of inheritance
between genomes,
arbitrary patterns of genetic inheritance can be directly expressed, such as
gene conversions,
multiple simultaneous recombinations,
and so on.
% Wilder: give forward ref to section on efficiency here, mention tree sequence
Thirdly, local trees that we generate for a particular position on the
genome can have nodes of any arity (number of child nodes), including ``unary''
nodes with a single child (see Section~\ref{sec-locally-unary-edges}).
% TODO should rephrase this from parent/child to ancestor/descendant?
Fourth, recombination is modelled in terms of
its inputs (parental genomes) and output (child genome) with associated inheritance intervals.
Thus we will have a child node (the recombinant; e.g.\ node \noderef{a}
Fig.~\ref{fig-arg-in-pedigree}) and two (or more; see Section~\ref{sec-ARG-simplification})
% TODO: check above if we want to refer to sec-ARG-simplification or some other!
parental nodes from which this child has inherited genome segments.
% TODO rephrase the above, it'll confuse people. Just need to reiterate the
% points about events, and maybe lead into the next section
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the processes of genetic inheritance
with recombination,
and is agnostic to the details of those processes.

@@

\section{Event ARGs}
\label{sec-eARG}
Genome ARGs are defined by
describing the details of genetic inheritance
between ancestor and descendant genomes.
In contrast, ARGs are classically defined
by describing the \emph{events} in the history of a sample of genomes.
The difference between gARGs and the prevalent views of eARGs
% what defines an ARG
are subtle and difficult to distinguish.
It is therefore useful to first survey some representative
recent definitions of
an ARG from the literature, to allow us to capture their common features
(see Appendix~\ref{sec-arg-history} for a full review).

@@

Whether we regard nodes as events or genomes is something of a
philosophical question, and in most senses the two are equivalent
(but see the discussion on ``unary nodes'' in Section~\ref{sec-locally-unary-edges}).
From a concrete data-structure encoding perspective, the only real
difference between an eARG and a gARG is that in an eARG we store recombination
\emph{breakpoints} at nodes of out-degree two (recombination nodes)
whereas in a gARG we store \emph{inheritance intervals} on edges.
A key advantage of associating inheritance intervals with
edges in a gARG is that it allows us to describe genetic inheritance events other
than simple crossovers, such as multiple recombinations and
gene conversions. By explicitly stating the exact regions of genome
that are transmitted along an edge, we generalise the representation
to encompass any process that leads to the differential inheritance of
genetic material along the genome.

>>>>

% TODO add a few sentences here on why defining nodes concretely as
% integers has significant advantages.
% 1) leads to direct link between the ARG and local trees in terms
%    of a well-defined combinatorial object (local trees section)
% 2) Definitions can be implemented directly in an efficient way
%    (computational efficiency section.
In many settings, nodes are dated, i.e.\
each node $u\in N$ is associated with a time $\tau_u$.
However, the topological relationships
defined by the directed graph structure is sufficient for
many applications, and we will assume that node dates are not
known here.
% This should be obvious, but seems like it needs to be said
In practical settings, we will wish to associate additional
metadata such as sample identifiers or quality-control metrics
with nodes; it is therefore best to think of the
integers used here in the definition of a node as an \emph{identifier},
with which arbitrary additional information can be associated.

@@

As illustrated in Fig.~\ref{fig-arg-in-pedigree},
the gARG for a given set of individuals is embedded in their pedigree.
The pedigree relationship between the diploid individuals is
shown, along with their paths
of genetic inheritance and constituent genomes (labelled
with lowercase alphabetical letters
rather than integer identifiers to avoid confusion with genomic
intervals).
Thus, individual $D_1$ is composed
of genomes \noderef{a} and \noderef{b}, which are inherited from its
two parents, $D_3$ and $D_4$. Each inherited genome may be the recombined product
of the genomes
% FIXME "contained within" is a poor phrase here, use "constituent" or something
contained within a parent.
In this example,
genome \noderef{b} was inherited directly from $D_4$'s genome \noderef{g} without
recombination, whereas
genome \noderef{a} is the recombinant product of
$D_2$'s genomes \noderef{e} and \noderef{f} crossing over at position 2.
Specifically, genome \noderef{a} inherited the (half-closed)
interval $[0, 2)$ from genome \noderef{e} and $[2, 10)$ from genome \noderef{f}.
These intervals are shown attached to the corresponding graph edges.
The figure shows the annotated pedigree with realised inheritance of genomes
between generations, the corresponding gARG, and finally the corresponding
sequence of local trees along the genome.
The local trees represent the genealogy of genomes \noderef{a}--\noderef{d},
which changes between local trees due to recombination.
The local trees span the three genome regions delineated
by the two recombination breakpoints that gave rise to these genomes.
Following the above definitions, corresponding data encoding for
this gARG is shown in Tab.~\ref{tab-gARG-data} (Appendix~\ref{sec-gARG-data}).

These definitions follow naturally from the outcome of events
associated with the inheritance of DNA between generations.
% FIXME this sentance doesn't follow well from the first, needs a
% different bridge before listing important properties.
% Maybe the "listicle" format here isn't right anyway.
As such,
there are some important consequences to these definitions.
Firstly, there is
no limitation on a mating system, ploidy, or age structure in a gARG.
By making the
basic unit an individual's genomes,
any pattern of inheritance among mixed ploidy (e.g.\ haplodiploid
species) can be accommodated.
Secondly, because we record the intervals of inheritance
between genomes,
arbitrary patterns of genetic inheritance can be directly expressed, such as
gene conversions,
multiple simultaneous recombinations,
and so on.
% Wilder: give forward ref to section on efficiency here, mention tree sequence
% TODO, should probably drop this point.
Thirdly, local trees that we generate for a particular position on the
genome can have nodes of any arity (number of child nodes), including ``unary''
nodes with a single child (see Section~\ref{sec-locally-unary-edges}).
% TODO should rephrase this from parent/child to ancestor/descendant?
Fourth, recombination is modelled in terms of
its inputs (parental genomes) and output (child genome) with associated inheritance intervals.
Thus we will have a child node (the recombinant; e.g.\ node \noderef{a}
Fig.~\ref{fig-arg-in-pedigree}) and two (or more; see Section~\ref{sec-ARG-simplification})
% TODO: check above if we want to refer to sec-ARG-simplification or some other!
parental nodes from which this child has inherited genome segments.
% TODO rephrase the above, it'll confuse people. Just need to reiterate the
% points about events, and maybe lead into the next section
Finally, gARGs are not defined in terms of the
coalescent with recombination (or any other) stochastic process.
Thus, the gARG encoding is focused on faithfully representing the
\emph{outcome} of the processes of genetic inheritance
with recombination,
and is agnostic to the details of those processes.

@@

\section{Event ARGs}
\label{sec-eARG}
% FIXME - this intro doesn't work at all now - we haven't defined what an
% eARG *is* before discussing it.
% Maybe best to skip this entirely, and make the quoted definitions with
% a thin intro the way to start the section.
Genome ARGs are defined by
describing the details of genetic inheritance
between ancestor and descendant genomes.
In contrast, ARGs are classically defined
by describing the \emph{events} in the history of a sample of genomes.
The difference between gARGs and the prevalent views of eARGs
% what defines an ARG
are subtle and difficult to distinguish.
It is therefore useful to first survey some representative
recent definitions of
an ARG from the literature, to allow us to capture their common features
(see Appendix~\ref{sec-arg-history} for a full review).

@@

% TODO reference forward to the cell ARG appendix. Ultimately
% every event corresponds to a particular genome in a particular
% individual or cell.
Whether we regard nodes as events or genomes is something of a
philosophical question, and in most senses the two are equivalent
(but see the discussion on ``unary nodes'' in Section~\ref{sec-locally-unary-edges}).
From a concrete data-structure encoding perspective, the only real
difference between an eARG and a gARG is that in an eARG we store recombination
\emph{breakpoints} at nodes of out-degree two (recombination nodes)
whereas in a gARG we store \emph{inheritance intervals} on edges.
A key advantage of associating inheritance intervals with
edges in a gARG is that it allows us to describe genetic inheritance events other
than simple crossovers, such as multiple recombinations and
gene conversions. By explicitly stating the exact regions of genome
that are transmitted along an edge, we generalise the representation
to encompass any process that leads to the differential inheritance of
genetic material along the genome.
